---
title: "Gecko"
author: "Rong Guang"
date: "2023-11-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Read in data

```{r}
#read data
library(tidyverse)
library(here)
gecko_full_vars <- read_csv("Dataframe.csv")
```
## Select variables for modelling

SELECT variables for modeling into object `gecko`. Variable names are mostly the original ones, except that:
    
    - all `()` are turned into `_`, e.g. "hydroxyl (alkyl)" turned into hydroxyl_alkyl
    
    - all spaces are turned into `_`, e.g. "aromatic(hydroxyl)" turned into aromatic_hydroxyl
    
    - two special cases are : now "C=C-C=O in non-aromatic ring" is "ccco", and  "C=C (non-aromatic)" is "cc",. 

```{r}
# SELECT variables into object `gecko`
gecko <- 
  gecko_full_vars |> 
  dplyr::select(
    id = index,           #A unique molecule index used in naming files
    MW,                   #The molecular weight of the molecule (g/mol).
    pSat_Pa,              #The saturation vapour pressure of the molecule calculated by COSMOtherm (Pa)
    NumOfAtoms,           #The number of atoms in the molecule
    NumOfC,               #The number of carbon atoms in the molecule
    NumOfO,               #The number of oxygen atoms in the molecul
    NumOfN,               #The number of nitrogen atoms in the molecule
    NumHBondDonors,       #“The number of hydrogen bond donors in the molecule, i.e. hydrogens bound to oxygen.”
    NumOfConf,            #The number of stable conformers found and successfully calculated by COSMOconf.
    NumOfConfUsed,        #The number of conformers used to calculate the thermodynamic properties.
    parentspecies,        #Either “decane”, “toluene”, “apin” for alpha-pinene or a combination of these connected by an underscore to indicate ambiguous                           #descent. In 243 cases, the parent species is “None” because it was not possible to retrieve it.
    
    cc =                  #The number of non-aromatic C=C bounds found in the molecule.
      "C=C (non-aromatic)", 
    
    ccco =               #The number of “C=C-C=O” structures found in non-aromatic rings in the molecule.
      "C=C-C=O in non-aromatic ring",
    
    hydroxyl_alkl =      #The number of the alkylic hydroxyl groups found in the molecule.
      "hydroxyl (alkyl)",
    
    aldehyde,            #The number of aldehyde groups in the molecule.
    ketone,              #The number of ketone groups in the molecule.
    
    carboxylic_acid =    #The number of carboxylic acid groups in the molecule.
      "carboxylic acid",
    ester,               #The number of ester groups in the molecule.
    
    ether_alicyclic =    #The number of alicyclic ester groups in the molecule.
      "ether (alicyclic)",
    nitrate,             #The number of alicyclic nitrate groups in the molecule
    nitro,               #The number of nitro ester groups in the molecule
    
    aromatic_hydroxyl =
    "aromatic hydroxyl", #The number of alicyclic aromatic hydroxyl groups in the molecule.
    
    carbonylperoxynitrate, #The number of carbonylperoxynitrate groups in the molecule.
    peroxide,            #The number of peroxide groups in the molecule
    hydroperoxide,       #The number of hydroperoxide groups in the molecule.
    carbonylperoxyacid,  #The number of carbonylperoxyacid groups found in the molecule
    nitroester           #The number of nitroester groups found in the molecule
  )
```

## Assign variable labels

```{r}
library(labelled)
var_label(gecko) <- 
  c(
    " A unique molecule index used in naming files",
    "The molecular weight of the molecule (g/mol)",
    "The saturation vapour pressure of the molecule calculated by COSMOtherm (Pa)",
    "The number of atoms in the molecule",
    "The number of carbon atoms in the molecule",
    "The number of oxygen atoms in the molecul",
    "The number of nitrogen atoms in the molecule",
    "The number of hydrogen bond donors in the molecule, i.e. hydrogens bound to oxygen.",
    "The number of stable conformers found and successfully calculated by COSMOconf.",
    "The number of conformers used to calculate the thermodynamic properties.",
    "Either “decane”, “toluene”, “apin” for alpha-pinene or a combination of these connected by an underscore to indicate ambiguousdescent. In 243 cases, the parent species is “None” because it was not possible to retrieve it.",
    "The number of non-aromatic C=C bounds found in the molecule.",
    "The number of “C=C-C=O” structures found in non-aromatic rings in the molecule.",
    "The number of the alkylic hydroxyl groups found in the molecule",
    "The number of aldehyde groups in the molecule.",
    "The number of ketone groups in the molecule.",
    "The number of carboxylic acid groups in the molecule.",
    "The number of ester groups in the molecule.",
    "The number of alicyclic ester groups in the molecule.",
    "The number of alicyclic nitrate groups in the molecule",
    "The number of nitro ester groups in the molecule",
    "The number of alicyclic aromatic hydroxyl groups in the molecule.",
    "The number of carbonylperoxynitrate groups in the molecule.",
    "The number of peroxide groups in the molecule",
    "The number of hydroperoxide groups in the molecule.",
    "The number of carbonylperoxyacid groups found in the molecule",
    "The number of nitroester groups found in the molecule"
  )
```

```{r}
# use example
var_label(gecko$carboxylic_acid)
```

## Varible type

```{r}
var.type <- gecko |> map(class) |> unlist() |> data.frame()
names(var.type) <- "var_type"
var.type
```

## Handle categorical variables

Find character variable

```{r}
var.type |> 
  filter(
    var_type == "character"
  )
```

Examine it

```{r}
gecko$parentspecies |> table()
```

One hot encoding:

```{r}
library(fastDummies)

#extract categorical variable 
category_var <- gecko[,c("id", "parentspecies")]

#one hot encoding
gecko_cat <- category_var |> fastDummies::dummy_cols(select_columns = "parentspecies")

#for convenience of visualization, numeric variables are filtered out:
gecko_num <- 
  gecko |> 
  dplyr::select(
    -parentspecies
  )
```


## Visualize numeric variables

### Box plot

```{r, fig.height= 5, fig.width= 8}
gecko_num[,-1] |> 
  pivot_longer(
    everything(),
    values_to = "values",
    names_to = "variables"
    ) |> 
  ggplot(
    aes(x = values)
  ) +
  geom_boxplot() +
  facet_wrap(
    ~variables, 
    scales = "free"
  )+
  theme_bw()
```

### Histogram

```{r, fig.height= 5, fig.width= 8}
gecko_num[,-1] |> 
  pivot_longer(
    everything(),
    values_to = "values",
    names_to = "variables"
    ) |> 
  ggplot(
    aes(x = values)
  ) +
  geom_histogram() +
  facet_wrap(
    ~variables, 
    scales = "free"
  )+
  theme_bw()
```

## Log/exp transformation for non-normal data

### NumOfConf and MW

The distribution of `NumOfConf` is skewed.

```{r}
NumOfConf.dist.a <- 
  gecko_num |> 
  ggplot(aes(x = NumOfConf^0.3)) +
  geom_qq()

NumOfConf.dist.b <- 
  gecko_num |> 
  ggplot(aes(sample = NumOfConf^0.3)) +
  geom_qq()
```

```{r}
gecko_num <- 
  gecko_num |> 
  mutate(NumOfConf_exp = NumOfConf^0.3)
```

```{r}
gecko_num |> 
  ggplot(aes(sample = NumOfConf^0.3)) +
  geom_qq()
```


### check

```{r}
gecko_num |> 
  ggplot(aes(x = NumOfConf |> log()))+
  geom_histogram(binwidth = 1)

```





```{r}
gecko$peroxide |> table()
var_label(gecko$hydroperoxide)
```

## handle missing values

There is no missing values in our data 

```{r fig.width=8, fig.height=4}
library(naniar)
vis_miss(gecko_num)
```

## Handle out-liers

```{r}
gecko_num_scaled <- scale(gecko_num[,-1]) |> data.frame()

gecko_num_scaled |> 
  mutate(id = gecko_num$id) |> 
  select(id, everything())
```

```{r}
gecko_num_scaled |> 
  filter(abs(MW) > 3.29) |> 
  select (MW)
```

MW; NumOfAtoms; NumOfConf; NumOfO

```{r}
gecko_num_scaled |>
  select(
    MW,
    NumOfAtoms,
    NumOfConf,
    NumOfO
  ) |> 
  apply(2, function(x)sum(abs(x)>3.29))
```

```{r}

```




